String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-08-30 2023-1-19/sections/introduction.tex
intro

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from such string constraints, especially when the counting and length bounds are large. For instance, none of the string solvers CVC5, Z3seq, Z3-Trau, Z3str3, Z3str3RE, and OSTRICH is capable of solving the following constraint within 300 seconds, and Z3 can not solve it within 120 seconds.
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c^+ \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three strings $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with a nonempty sequence of $c$'s, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in both $(\Sigma \setminus a)^{\{1, 60\}}$ and $(\Sigma \setminus b)^{\{1, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c^+$ requires that $x$ has to end with $c$.

A typical way for string constraint solvers to deal with regular expressions with counting is to unfold them into those \emph{without} counting using the concatenation operator. For instance, $a^{\{1, 4\}}$ is unfolded into $a(\varepsilon + a + aa + aaa)$ and $a^{\{2,\infty\}}$ is unfolded into $aaa^{*}$. Since the unfolding incurs an exponential blow-up on the sizes of constraints (assuming that the counting in string constraints are encoded in binary), the unfolding penalizes the performance of the solvers, especially when the bounds are large.

@@

\medskip
\noindent 
\emph{Contribution.} In this work, we focus on the class of string constraints with regular membership and string-length function, where the counting operators may occur (called RECL for brevity). We make the following contributions: 
\vspace{-0.5mm}
\begin{itemize}
  \item We propose an automata-theoretical approach for solving RECL constraints. 
  Our main idea is to represent the counting operators by registers in cost-enriched finite automata (CEFA, see Section~\ref{sec:automaton} for the definition), instead of unfolding them explicitly. The satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA w.r.t. a linear integer arithmetic (LIA), which is then transformed to the satisfiability of existential LIA formula tackled by the off-the-shelf SMT solver.
  % Moreover, the string-length function can also be modeled by register in CEFA. Then the satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA with respect to a linear integer arithmetic (LIA) formula whose free variables are registers. According to the results from~\cite{atva2020}, an LIA formula can be computed to represent the potential values of registers in CEFA. 
  % Thus, the nonemptiness of CEFA w.r.t. LIA formulas can be reduced to the satisfiability of LIA formulas, which is then tackled by off-the-shelf SMT solvers.
  \item We propose simplification techniques to reduce the state space and the number of transitions in CEFA, to achieve better performance. It is beneficial since the sizes of LIA formulas depend much on the sizes of CEFA.
  \item Combined with the simplification techniques mentioned above, the register representation of regex-counting and string-length in CEFA entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting in a practical string solver called $\ostrichrecl$. 
  \item Finally, we utilize a collection of benchmark suites comprising 48,843 instances in total to evaluate the performance of our tool. The result shows that $\ostrichrecl$ solves the most instances among all the solvers, with comparable speed (See Table~\ref{tab:results_regcol}). In addition, we carry out some detailed experiments, including the evaluation of simplification techniques and the experiment of large-counting benchmarks. The result shows that the simplification techniques are effective in practice (See Table~\ref{tab:results_simp}), and $\ostrichrecl$ performs much better than other solvers on the large-counting benchmark (See Fig~\ref{fig:table_large_count}).
  % The experiment results show that $\ostrichrecl$ can solve 48,698 out of 48,843 instances, while Z3str3RE, the state-of-the-art solver on RECL constraints, solves only 47,607 instances (where the timeout period is set as 60 seconds). Moreover, the speed of $\ostrichrecl$ is comparable to that of Z3str3RE. In comparison, CVC5, Z3seq, Z3-Trau, and Z3str3 solve only 44,704, 46,426, 24,121, and 35,866 instances respectively. (See Table~\ref{tab:results_regcol}-\ref{tab:results_summary} for more details.)
\end{itemize}
\vspace{-0.5mm}
% We propose an automata-theoretical approach for solving RECL constraints. Our main idea is to represent the counting operators by registers in automata (as much as possible) instead of unfolding them explicitly. Our approach is built upon cost-enriched finite automata (CEFA), proposed in~\cite{atva2020} to solve string constraints involving the integer data type. Moreover, the string-length function can also be modeled by register in CEFA. Then the satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA with respect to a linear integer arithmetic (LIA) formula whose free variables are registers of CEFA.

@@

\medskip
\noindent
\emph{Related work.} 
We discuss more related work about regexes with counting in programming languages.  
The determinism of regexes with counting was investigated in~\cite{GGM12,CL15}. Counting-set automata were proposed in~\cite{redos_lenka,HS+23} to quickly match a subclass of regexes with counting. Moreover, a variant of nondeterministic counter automata, called bit vector automata, was proposed recently in \cite{GKM23} to enable fast matching of a more expressive class of regexes with counting.   Nevertheless, the nonemptiness problem of these automata was not considered, and it is unclear whether these automata models can be used for solving string constraints with regex-counting and string-length.
Real-world regexes in programming languages include features beyond classical regular expressions, e.g., the greedy/lazy Kleene star, capturing groups, and back references. Real-world regexes have been addressed in symbolic execution of Javascript programs~\cite{LMK19} and string constraint solving~\cite{CF+22}. Nevertheless, the counting operators are still unfolded explicitly in~\cite{CF+22}.

>>>>

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from such string constraints, especially when the counting and length bounds are large. For instance, none of the string solvers CVC5, Z3seq, Z3-Trau, Z3str3, Z3str3RE, and OSTRICH is capable of solving the following constraint within 120 seconds.
%, and Z3 can not solve it within 120 seconds.
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c^+ \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three strings $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with a nonempty sequence of $c$'s, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in both $(\Sigma \setminus a)^{\{1, 60\}}$ and $(\Sigma \setminus b)^{\{1, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c^+$ requires that $x$ has to end with $c$.

A typical way for string constraint solvers to deal with regular expressions with counting is to unfold them into those \emph{without} counting using the concatenation operator. For instance, $a^{\{1, 4\}}$ is unfolded into $a(\varepsilon + a + aa + aaa)$ and $a^{\{2,\infty\}}$ is unfolded into $aaa^{*}$. Since the unfolding incurs an exponential blow-up on the sizes of constraints (assuming that the counting in string constraints are encoded in binary), the unfolding penalizes the performance of the solvers, especially when the length bounds are also available.

@@

\medskip
\noindent 
\emph{Contribution.} In this work, we focus on the class of string constraints with regular membership and string-length function, where the counting operators may occur (called RECL for brevity). We make the following contributions.
\vspace{-0.5mm}
\begin{itemize}
  \item We propose an automata-theoretical approach for solving RECL constraints. 
  Our main idea is to represent the counting operators by cost registers in cost-enriched finite automata (CEFA, see Section~\ref{sec:automaton} for the definition), instead of unfolding them explicitly. The string-length function is modeled by cost registers as well. The satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA w.r.t. a linear integer arithmetic (LIA) formula. According to the results from~\cite{atva2020}, an LIA formula can be computed to represent the potential values of registers in CEFA.
Thus, the nonemptiness of CEFA w.r.t. LIA formulas can be reduced to the satisfiability of LIA formulas, which is then tackled by off-the-shelf SMT solvers.
%
  % Moreover, the string-length function can also be modeled by register in CEFA. Then the satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA with respect to a linear integer arithmetic (LIA) formula whose free variables are registers. According to the results from~\cite{atva2020}, an LIA formula can be computed to represent the potential values of registers in CEFA. 
  % Thus, the nonemptiness of CEFA w.r.t. LIA formulas can be reduced to the satisfiability of LIA formulas, which is then tackled by off-the-shelf SMT solvers.
  \item We propose simplification techniques to reduce the state space and the number of transitions in CEFA, to achieve better performance.
%  , since the sizes of LIA formulas depend much on the sizes of CEFA.
  \item Combined with the simplification techniques mentioned above, the register representation of regex-counting and string-length in CEFA entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting in a string solver called $\ostrichrecl$. 
 %
  \item Finally, we utilize a collection of benchmark suites comprising 48,843 instances in total to evaluate the performance of $\ostrichrecl$. The experiment results show that $\ostrichrecl$ solves the RECL constraints more efficiently than the state-of-the-art string solvers, especially when the counting and length bounds are large (see Table~\ref{tab:results_regcol} and Figure~\ref{fig:table_large_count}). For instance, on 1,873 benchmark instances where the counting bounds are greater than or equal to $50$ and the string lengths are required to be beyond $200$, $\ostrichrecl$ solves at least $278$ more instances than the other solvers, while spending only half or less time per instance on average. 
%  For instance,  the most number of instances than the other solvers, with comparable or faster speed (See Table~\ref{tab:results_regcol}). Moreover, when the counting and length bounds are large, $\ostrichrecl$ solves much more instances than the other solvers, while using much less time.

%In addition, we carry out some detailed experiments, including the evaluation of simplification techniques and the experiment of large-counting benchmarks. The result shows that the simplification techniques are effective in practice (See Table~\ref{tab:results_simp}), and $\ostrichrecl$ performs much better than other solvers on the large-counting benchmark (See Fig~\ref{fig:table_large_count}).

% The experiment results show that $\ostrichrecl$ can solve 48,698 out of 48,843 instances, while Z3str3RE, the state-of-the-art solver on RECL constraints, solves only 47,607 instances (where the timeout period is set as 60 seconds). Moreover, the speed of $\ostrichrecl$ is comparable to that of Z3str3RE. In comparison, CVC5, Z3seq, Z3-Trau, and Z3str3 solve only 44,704, 46,426, 24,121, and 35,866 instances respectively. (See Table~\ref{tab:results_regcol}-\ref{tab:results_summary} for more details.)
\end{itemize}
\vspace{-0.5mm}
% We propose an automata-theoretical approach for solving RECL constraints. Our main idea is to represent the counting operators by registers in automata (as much as possible) instead of unfolding them explicitly. Our approach is built upon cost-enriched finite automata (CEFA), proposed in~\cite{atva2020} to solve string constraints involving the integer data type. Moreover, the string-length function can also be modeled by register in CEFA. Then the satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA with respect to a linear integer arithmetic (LIA) formula whose free variables are registers of CEFA.

@@

\medskip
\noindent
\emph{Related work.} 
We discuss more related work about regexes with counting in programming languages.  
The determinism of regexes with counting was investigated in~\cite{GGM12,CL15}. Counting-set automata were proposed in~\cite{redos_lenka,HS+23} to quickly match a subclass of regexes with counting. Moreover, a variant of nondeterministic counter automata, called bit vector automata, was proposed recently in \cite{GKM23} to enable fast matching of a more expressive class of regexes with counting.   Nevertheless, the nonemptiness problem of these automata was not considered, and it is unclear whether these automata models can be used for solving string constraints with regex-counting and string-length.
Real-world regexes in programming languages include features beyond classical regular expressions, e.g., the greedy/lazy Kleene star, capturing groups, and back references. Real-world regexes have been addressed in symbolic execution of Javascript programs~\cite{LMK19} and string constraint solving~\cite{CF+22}. Nevertheless, the counting operators are still unfolded explicitly in~\cite{CF+22}.  Solving string constraints involving lengths has also been investigated by various researchers, see e.g. \cite{WC+18,atva2020,z3str3re}.
